Move Schedules: Fast persistence computations in coarse dynamic settings\thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}

### 2021-09-28 move_schedules.tex
Much better reduction of lcs edit distance -> permutation edit distance

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches.
  Since it is unknown a priori at which times  the persistent homology will change, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagrams at each one of them  independently. An alternative approach is to construct a homotopy between two contiguous time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. 
  The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. 
  As one can see from the figure, if one needs to simulate the homotopy exactly, then the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is sparse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

@@

\subsubsection{The Discrete Setting} 
Since moves do not maintain a valid decomposition during intermediate steps, it is worth asking if  the vineyard bounds can be improved by 
relaxing the goal of simulating persistence across  $F : K \times [0,1] \to \mathbb{R}$. 
In the case of vineyards, the bound is clear: 
%whether assuming a homotopy between a pair of filtrations $(K_0, K_1)$ or not, 
if $\kappa = K_\tau(K_0, K_1)$, then $\Omega(\kappa)$ adjacent transpositions are required to transform $K_0 \mapsto K_1$. 
In contrast, if we allow move operations in any order, we have $O(m)$ as a trivial upper bound: simply move each simplex in $K_0$ into its position in the filtration given by $K_1$ in the order given by the latter. However, it's not immediately clear whether this bound is tight. To establish this formally, we require a few definitions.

Let $S_m$ denote the symmetric group  of   permutations of  $[m]$ under function composition $\circ$.
Given two fixed permutations $p, q \in S_m$ and a set $\Sigma \subseteq S_m$,  the following 
are common problems in the literature:
\begin{enumerate}
	\item Find an ordered sequence of permutations $s_1, s_2, \dots, s_d \in \Sigma$ whose composition transforms $p$ into $q$:
	\begin{displaymath}
		s_d \circ \cdots \circ s_2 \circ s_1 \circ p = q
		% p \sbullet s_1 \sbullet s_2 \sbullet \cdots \sbullet s_d = q
	\end{displaymath}
	\item Find a sequence satisfying (1) of minimal length ($d$)
	\item Find the minimal length $d$, referred to as the \emph{distance} between $p$ and $q$ with respect to $\Sigma$. 
\end{enumerate}
A sequence $S = \{s_1, s_2, \dots, s_d\} \subseteq \Sigma \subseteq S_m$ of operations mapping $p \mapsto q$ is sometimes called a \emph{sorting of $p$}.
In the context of performing updates to the $R = D V$ decomposition using moves,  such sequences will form schedules. 
There are typically many ways to solve (1), and any solution to (2) necessarily solves (1) and (3). When $p, q$ are interpreted as strings, distances defined with respect to a fixed $\Sigma \subseteq S_m$ are commonly referred to as \emph{edit distances}~\cite{labarre2013lower}, which are denoted as $d_\Sigma(\cdot, \cdot)$. 
%Note that here we describe distances between permutations (e.g. $p,q \in S_n$) using operations that can themselves be expressed as permutations (e.g. $s \in S \subseteq S_n$), and that 
The choice of $\Sigma$ defines the universe of operations and thus the type of edit distance being measured---otherwise if $\Sigma = S_m$, then $d_\Sigma(p, q) = 1$ for any $p\neq  q \in S_m$.

@@

Here we consider sorting by moves.
In terms of permutations, 
moving  $i$ to $j$ in 
$[m]$, for $i < j$,
corresponds to 
\[
\scalefont{1.0}{
\setcounter{MaxMatrixCols}{20}
\left (
\begin{array}{*3c | *5c | *{3}c } 
\cline{4-8} \rule{0pt}{2ex}
1 & \cdots & i -1 & i & i + 1 & \cdots & j - 1 & j & j +1 &  \cdots & m \\
\cline{4-8}
\cline{4-8}\rule{0pt}{2ex}
1 & \cdots & i -1 & i + 1 & \cdots & j - 1 & j & i & j +1 &  \cdots & m \\
\cline{4-8}
\end{array}
\right ) 
}
\]
while moving $j$ to $i$ is the inverse permutation.
%The complexity of sorting by moves %is related to the well-known \emph{Levenshtein distance}~\cite{bar2004approximating}, which is the minimum number of character insertions, deletions, and substitutions needed to transform one string into another. 
%Through a dynamic programming type approach, the Wagner-Fischer algorithm can compute the Levenstein distance in $O(m^2)$, and there are several sub-quadratic approximations~\cite{bar2004approximating}. 
%is related to the \emph{LCS} distance, which is the minimum number of character insertions or deletions needed to transform one string into another. 
%In terms of hardness, there is evidence to suggest that the lower bound on computing Levenshtein distance in full generality is near quadratic time. 
%
%Like the \emph{LCS} distance, 
Sorting by move operations 
can be interpreted as finding a minimal sequence of edits where the only operations allowed are insertions and deletions, where each insertion/deletion operation is coupled. 
This is exactly the LCS (Longest Common Subsequence) distance. 
If these operations each take uniform cost and the universe of strings is restricted to $S_m$, the corresponding edit distance has a special name: the \emph{permutation edit distance}. We obtain the following lower bound size of a sorting (i.e., a schedule) using moves.

\begin{proposition}[Schedule size]
Given filtrations $K_0$ and $ K_1$ of size $m$, and an order preserving bijection $f: K_0 \to K_1$, 
then the smallest move schedule $S^\ast$ is of size $\Omega(d)$, where:
$$ d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert $$ 
Moreover, any size $d$ schedule $S$ can be computed in $O(m \log \log m)$ time. 
\end{proposition}
\begin{proof}
Recall our definition of  edit distance given above, 
depending on the choice of $S \subseteq S_m$ of allowable edit operations, and that in order for any edit distance to be symmetric, if $s \in S$ then $s^{-1} \in S$. This implies that $d_S(p,q) = d_S(p^{-1}, q)$ for any choice of  $p,q \in S_m$. 
Moreover, edit distances are \emph{left-invariant}, i.e.
\[
d_S(p,q) = d_S(r \circ p, r \circ q) \quad \text{ for all } p,q,r \in S_m
\]
Conceptually, left-invariance implies that the edit distance between any pair of permutations $p,q$ is invariant under an arbitrary relabeling $p,q$--as long as the relabeling is consistent. Thus, the following identity always holds: 
$$ d_S(p,q) = d_S(\iota, p^{-1} \circ q) = d_S(q^{-1} \circ p, \iota) $$
where $\iota = [m]$, the identity permutation. Suppose we are given permutations $p = K_0, q = K_1$, and we seek to compute $\mathrm{LCS}(p, q)$. Consider the permutation $p' = q^{-1} \circ p$. Since the LCS distance is a valid edit distance, if $\lvert \mathrm{LCS}(p, q) \rvert = k$, then $\lvert \mathrm{LCS}(p', \iota) \rvert = k$ as well. Notice that $\iota$ is strictly increasing and that any common subsequence $\iota$ has with $p'$ must also be strictly increasing. Thus, the problem of computing $\mathrm{LCS}(p, q)$ reduces to the problem of computing the \emph{longest increasing subsequence} (LIS) of $p'$, which can done in $O( m \log \log m)$ time using van Emde Boas trees~\cite{bespamyatnikh2000enumerating}. The optimality of $d$ follows from the optimality of the well-studied LCS problem~\cite{kumar1987linear}. 
\end{proof}

@@

 Before introducing the algorithm, we require a few definitions. Recall that a \emph{sorting} $\mathcal{S}$ with respect to two permutations $p, q \in S_m$ is an ordered sequence of permutations $(s_1, s_2, \dots, s_d)$ whose product under composition with $p$ yields $q$, i.e.
 $$ q = s_d \circ \dots s_1 \circ p $$
By definition, a subsequence $\mathcal{L}$ of symbols common to both $p$ and $q$ satisfies:
 $$ p^{-1}(\sigma) < p^{-1}(\tau) \implies q^{-1}(\sigma) < q^{-1}(\tau) \quad \forall \sigma, \tau \in \mathcal{L} $$
 where $p^{-1}(\sigma)$ (resp. $q^{-1}(\sigma)$) denotes the position of $\sigma$ in $p$ (resp. $q$). Thus, obtaining a sorting $p \mapsto q$ of size $d = m - \lvert \mathcal{L} \rvert$ reduces to applying a sequence of edit operations to symbols in the complement of $\mathcal{L}$. To ensure only $2d$ such operations are required, insertion operations must increase the size of $\mathcal{L}$ monotonically. 
 Suppose further we require these edit operations be coupled in the sense that for every deletion operation $\mathrm{del}_i$ of the symbol at position $i$, we compose with an insertion operation $\mathrm{ins}_j$ at position $j$. In this sense, we call the permutation a \emph{move} permutation if it a composition of insertion and deletion operations: 
 Similarly, we define a move permutation $s \in S_m$ as a \emph{valid} operation with respect to a fixed pair $p, q \in S_m$ if:
 \begin{equation}\label{eq:valid_move}
 	\lvert \mathrm{LCS}(s \circ p, q) \rvert = \lvert \mathrm{LCS}(p,q) \rvert + 1
  The problem of constructing a sorting $\mathcal{S}$ of size $d$ thus reduces to the problem of choosing a sequence of $d$ valid moves. We call such a sorting a valid sorting. To do this efficiently, we need a means of choosing a permutation $s \in S_m$ that satisfies equation~\ref{eq:valid_move}. 
  Let $\mathcal{T}$ denote a ordered set-like data structure that supports the following operations on elements $\sigma \in M$ from the set $M = \{0, 1, \dots, m+1\}$:
  	\item $\mathcal{T} \cup \sigma$---inserts $\sigma$ into $\mathcal{T}$,
  	\item $\mathcal{T} \setminus \sigma$---removes $\sigma$ from $\mathcal{T}$,
  	\item $\mathcal{T}_{\mathrm{succ}}(\sigma)$---obtain the successor of $\sigma$ in $\mathcal{T}$, if it exists, otherwise return $m+1$
  	\item $\mathcal{T}_{\mathrm{pred}}(\sigma)$---obtain the predecessor of $\sigma$ in $\mathcal{T}$, if it exists, otherwise return $0$
  Given such a structure $\mathcal{T}$, an arbitrary valid sorting can be constructed by repeatedly querying and updating information about the $\mathrm{LCS}$ in $\mathcal{T}$. To see this, suppose $\mathcal{T}$ contains all of the symbols in the current $\mathrm{LCS}$ between two permutations $p$ and $q$. By definition of the LCS, we have:
 \begin{equation}\label{eq:valid_move_ineq}
 	p^{-1}(\mathcal{T}_{\mathrm{pred}}(\sigma)) < p^{-1}(\sigma) < p^{-1}(\mathcal{T}_{\mathrm{succ}}(\sigma))
  for every $\sigma \in \mathcal{T}$. Now, suppose we choose a symbol $\sigma \notin \mathcal{T}$. If $p^{-1}(\sigma) < p^{-1}(\mathcal{T}_{\mathrm{pred}}(\sigma))$, then we must move $\sigma$ to the right in $p$ such that equation~\ref{eq:valid_move_ineq} holds. Similarly, if  $p^{-1}(\mathcal{T}_{\mathrm{succ}}(\sigma)) < p^{-1}(\sigma)$, then we must move $\sigma$ left in $p$ to increase the size of the LCS. Assuming the structure $\mathcal{T}$ supports all of the above operations in $O(\log m)$ time, we easily deduce a $O(d m\log m)$ algorithm for obtaining a valid sorting, which is shown in pseudocode in Algorithm~\ref{alg:sorting}.   
% one can incrementally build a valid sorting of size $d$ by arbitrarily choosing valid move operations . all elements in $\mathrm{LCS}(p,q)$ into $\mathcal{T}$
%  incrementally updating the inverse permutation $p^{-1}$ of $p$. Note that   
 % Further, let $\mathcal{I}_p$ denote an array of length $m = \lvert p \rvert$ whose values $\mathcal{I}(\sigma)$ store the position of the element $\sigma$ in $p$. Assume also that $\mathcal{I}_p()$  
 %for every pair of elements $\sigma, \tau \in \mathcal{L}$, if $p_{\sigma} < p_{\tau}$ where $p_{\sigma}$ denotes the the position of $\sigma$ in $p$, then $q_{\sigma} < q_{\tau}$ must also be true. 
 %under certain insertion and deletion operations.

@@

\subsubsection{Proxy objective}\label{sec:proxy_objective}
When $d \sim O(m)$, the quadratic nature of the aforementioned greedy minimization is too expensive. In order for Algorithm~\ref{alg:schedule} to be considered as an efficient alternative to vineyards---which requires $O(m \log m + k)$ preprocessing time to compute $k \sim O(m^2)$ transpositions---a solution which is subquadratic in $m$ is desirable.  %In what follows, we present such an objective, showing its relation to the problem at hand as well its use in other known computational problems. 
% We show empirical results demonstrating its practical effectiveness as a heuristic in section~\ref{sec:results}.

@@

\section{Conclusion and Future Work}\label{sec:conclusion}
In conclusion, we presented a scheduling algorithm for efficiently updating a decomposition in sparse dynamic settings. Our approach is simple, easy to implement, and fully general: it does not depend on the geometry of underlying space nor the choice of triangulation. Moreover, we supplied efficient algorithms for our scheduling strategy, provided tight bounds where applicable, and demonstrated our algorithms performance with several real world use cases.

>>>>

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches.
  Since it is unknown a priori at which times  the persistent homology will change, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagrams at each one of them  independently. An alternative approach is to construct a homotopy between two contiguous time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. 
  The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. 
  As one can see from the figure, if one needs to simulate the homotopy exactly, then the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

@@

\subsubsection{The Discrete Setting} 
At the finest scale, using move operations simulate persistence across $F : K \times [0,1] \to \mathbb{R}$ reduces to vineyards (i.e. when the moves themselves are adjacent transpositions).
The benefit of move operations is that one need not maintain a valid decomposition during the movement; thus, it is worth investigating whether the vineyard bounds can be improved by relaxing the assumption that one wants to simulate persistence across $F$ continuously. 
In the case of vineyards, the bound is clear: 
if $\kappa = K_\tau(K_0, K_1)$, then $\Omega(\kappa)$ adjacent transpositions are required to transform $K_0 \mapsto K_1$. 
In contrast, if we allow move operations in any order, we have $O(m)$ as a trivial upper bound: simply move each simplex in $K_0$ into its position in the filtration given by $K_1$ in the order given by the latter. However, it's not immediately clear whether this bound is tight. To establish this formally, we require a few definitions.

Let $S_m$ denote the symmetric group  of   permutations of  $[m]$ under function composition $\circ$.
Given two fixed permutations $p, q \in S_m$ and a set $\Sigma \subseteq S_m$,  the following 
are common problems in the literature:
\begin{enumerate}
	\item Find an ordered sequence of permutations $s_1, s_2, \dots, s_d \in \Sigma$ whose composition transforms $p$ into $q$:
	\begin{displaymath}
		s_d \circ \cdots \circ s_2 \circ s_1 \circ p = q
		% p \sbullet s_1 \sbullet s_2 \sbullet \cdots \sbullet s_d = q
	\end{displaymath}
	\item Find a sequence satisfying (1) of minimal length ($d$)
	\item Find the minimal length $d$, referred to as the \emph{distance} between $p$ and $q$ with respect to $\Sigma$. 
\end{enumerate}
A sequence $S = \{s_1, s_2, \dots, s_d\} \subseteq \Sigma \subseteq S_m$ of operations mapping $p \mapsto q$ is sometimes called a \emph{sorting of $p$}.
In the context of performing updates to the $R = D V$ decomposition using moves,  such sequences will form schedules. 
There are typically many ways to solve (1), and any solution to (2) necessarily solves (1) and (3). When $p, q$ are interpreted as strings, distances defined with respect to a fixed $\Sigma \subseteq S_m$ are commonly referred to as \emph{edit distances}~\cite{labarre2013lower}, which are denoted as $d_\Sigma(\cdot, \cdot)$, and the subsequent elementary operations on the strings are called \emph{edit operations}. 
%Note that here we describe distances between permutations (e.g. $p,q \in S_n$) using operations that can themselves be expressed as permutations (e.g. $s \in S \subseteq S_n$), and that 
The choice of $\Sigma$ defines the universe of operations and thus the type of edit distance being measured---otherwise if $\Sigma = S_m$, then $d_\Sigma(p, q) = 1$ for any $p\neq  q \in S_m$.

@@

Here we consider sorting by moves. In terms of permutations, a permutation $m_{ij}$ that moves $i$ to $j$ in $[m]$, for $i < j$, corresponds to: 
\[
\scalefont{1.0}{
\setcounter{MaxMatrixCols}{20}
\left (
\begin{array}{*3c | *5c | *{3}c } 
\cline{4-8} \rule{0pt}{2ex}
1 & \cdots & i -1 & i & i + 1 & \cdots & j - 1 & j & j +1 &  \cdots & m \\
\cline{4-8}
\cline{4-8}\rule{0pt}{2ex}
1 & \cdots & i -1 & i + 1 & \cdots & j - 1 & j & i & j +1 &  \cdots & m \\
\cline{4-8}
\end{array}
\right ) 
}
\]
In the context of edit operations, observe that a move operation can be interpreted the composition of deletion operation followed by an insertion operation: 
where $\mathrm{del}_i$ denotes the edit operation that deletes the character at position $i$ in some string $p$ and $\mathrm{ins}_j$ similarly inserts the same character at position $j$. Thus, sorting by move operations can be interpreted as finding a minimal sequence of edits where the only operations allowed are insertions and deletions, where each insertion/deletion operation is coupled.
This is exactly the LCS (Longest Common Subsequence) distance. It is well known~\cite{bergroth2000survey} that the LCS distance between two strings $p, q$ of sizes $m$ and $n$, respectively, is: 
	\mathrm{d}_{\mathrm{lcs}}(p,q) = m + n - 2\lvert LCS(p, q)\rvert
Intuitively, in the context of strings, given $\mathcal{L} = \mathrm{LCS}(p,q)$ one may transform $p \mapsto q$ by simply deleting the symbols from $p \setminus \mathcal{L}$ and then inserting symbols from $q \setminus \mathcal{L}$.
In the context of permutations, these operations are coupled, thus to ensure only $2d$ such edit operations are used to transform $p \mapsto q$, each move operations must increase the size of $\mathcal{L}$ monotonically. 
% while moving $j$ to $i$ is the inverse permutation.
%The complexity of sorting by moves %is related to the well-known \emph{Levenshtein distance}~\cite{bar2004approximating}, which is the minimum number of character insertions, deletions, and substitutions needed to transform one string into another. 
%Through a dynamic programming type approach, the Wagner-Fischer algorithm can compute the Levenstein distance in $O(m^2)$, and there are several sub-quadratic approximations~\cite{bar2004approximating}. 
%is related to the \emph{LCS} distance, which is the minimum number of character insertions or deletions needed to transform one string into another. 
%In terms of hardness, there is evidence to suggest that the lower bound on computing Levenshtein distance in full generality is near quadratic time. 
%Like the \emph{LCS} distance, 
Thus, by reducing the problem to the \emph{permutation edit distance}, we obtain the following lower bound on the minimum size of a sorting (i.e., a schedule) using moves. 
\begin{proposition}[Schedule size]
Given filtrations $K_0$ and $ K_1$ of size $m$, and an order preserving bijection $f: K_0 \to K_1$, 
then the smallest move schedule $S^\ast$ is of size $\Omega(d)$, where:
$$ d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert $$ 
Moreover, any size $d$ schedule $S$ can be computed in $O(m \log \log m)$ time. 
\end{proposition}
\begin{proof}
Recall our definition of  edit distance given above, 
depending on the choice of $S \subseteq S_m$ of allowable edit operations, and that in order for any edit distance to be symmetric, if $s \in S$ then $s^{-1} \in S$. This implies that $d_S(p,q) = d_S(p^{-1}, q)$ for any choice of  $p,q \in S_m$. 
Moreover, edit distances are \emph{left-invariant}, i.e.
\[
d_S(p,q) = d_S(r \circ p, r \circ q) \quad \text{ for all } p,q,r \in S_m
\]
Conceptually, left-invariance implies that the edit distance between any pair of permutations $p,q$ is invariant under an arbitrary relabeling $p,q$--as long as the relabeling is consistent. Thus, the following identity always holds: 
$$ d_S(p,q) = d_S(\iota, p^{-1} \circ q) = d_S(q^{-1} \circ p, \iota) $$
where $\iota = [m]$, the identity permutation. Suppose we are given permutations $p = K_0, q = K_1$, and we seek to compute $\mathrm{LCS}(p, q)$. Consider the permutation $p' = q^{-1} \circ p$. Since the LCS distance is a valid edit distance, if $\lvert \mathrm{LCS}(p, q) \rvert = k$, then $\lvert \mathrm{LCS}(p', \iota) \rvert = k$ as well. Notice that $\iota$ is strictly increasing and that any common subsequence $\iota$ has with $p'$ must also be strictly increasing. Thus, the problem of computing $\mathrm{LCS}(p, q)$ reduces to the problem of computing the \emph{longest increasing subsequence} (LIS) of $p'$, which can done in $O( m \log \log m)$ time using van Emde Boas trees~\cite{bespamyatnikh2000enumerating}. The optimality of $d$ follows from the optimality of the well-studied LCS problem~\cite{kumar1987linear}. 
\end{proof}

@@

 Before introducing the algorithm, we require a few definitions. Recall that a \emph{sorting} $\mathcal{S}$ with respect to two permutations $p, q \in S_m$ is an ordered sequence of permutations $(s_1, s_2, \dots, s_d)$ whose product under composition with $p$ yields $q$, i.e.
 $$ q = s_d \circ \dots s_1 \circ p $$
By definition, a subsequence $\mathcal{L}$ of symbols common to both $p$ and $q$ satisfies:
 $$ p^{-1}(\sigma) < p^{-1}(\tau) \implies q^{-1}(\sigma) < q^{-1}(\tau) \quad \forall \sigma, \tau \in \mathcal{L} $$
 where $p^{-1}(\sigma)$ (resp. $q^{-1}(\sigma)$) denotes the position of $\sigma$ in $p$ (resp. $q$). Thus, obtaining a sorting $p \mapsto q$ of size $d = m - \lvert \mathcal{L} \rvert$ reduces to applying a sequence of move operations to symbols in the complement of $\mathcal{L}$. 
 Formally, we define a move permutation $s \in S_m$ as a \emph{valid} operation with respect to a fixed pair $p, q \in S_m$ if:
 \begin{equation}\label{eq:valid_move}
 	\lvert \mathrm{LCS}(s \circ p, q) \rvert = \lvert \mathrm{LCS}(p,q) \rvert + 1
 \end{equation}
  The problem of constructing a sorting $\mathcal{S}$ of size $d$ thus reduces to the problem of choosing a sequence of $d$ valid moves. We call such a sorting a valid sorting. To do this efficiently, we need a means of choosing a permutation $s \in S_m$ that satisfies equation~\ref{eq:valid_move}. 
  Let $\mathcal{T}$ denote a ordered set-like data structure that supports the following operations on elements $\sigma \in M$ from the set $M = \{0, 1, \dots, m+1\}$:
  \begin{enumerate}
  	\item $\mathcal{T} \cup \sigma$---inserts $\sigma$ into $\mathcal{T}$,
  	\item $\mathcal{T} \setminus \sigma$---removes $\sigma$ from $\mathcal{T}$,
  	\item $\mathcal{T}_{\mathrm{succ}}(\sigma)$---obtain the successor of $\sigma$ in $\mathcal{T}$, if it exists, otherwise return $m+1$
  	\item $\mathcal{T}_{\mathrm{pred}}(\sigma)$---obtain the predecessor of $\sigma$ in $\mathcal{T}$, if it exists, otherwise return $0$
  \end{enumerate}
  Given such a structure $\mathcal{T}$, an arbitrary valid sorting can be constructed by repeatedly querying and updating information about the $\mathrm{LCS}$ in $\mathcal{T}$. To see this, suppose $\mathcal{T}$ contains all of the symbols in the current $\mathrm{LCS}$ between two permutations $p$ and $q$. By definition of the LCS, we have:
 \begin{equation}\label{eq:valid_move_ineq}
 	p^{-1}(\mathcal{T}_{\mathrm{pred}}(\sigma)) < p^{-1}(\sigma) < p^{-1}(\mathcal{T}_{\mathrm{succ}}(\sigma))
 \end{equation}
  for every $\sigma \in \mathcal{T}$. Now, suppose we choose a symbol $\sigma \notin \mathcal{T}$. If $p^{-1}(\sigma) < p^{-1}(\mathcal{T}_{\mathrm{pred}}(\sigma))$, then we must move $\sigma$ to the right in $p$ such that equation~\ref{eq:valid_move_ineq} holds. Similarly, if  $p^{-1}(\mathcal{T}_{\mathrm{succ}}(\sigma)) < p^{-1}(\sigma)$, then we must move $\sigma$ left in $p$ to increase the size of the LCS. Assuming the structure $\mathcal{T}$ supports all of the above operations in $O(\log m)$ time, we easily deduce a $O(d m\log m)$ algorithm for obtaining a valid sorting, which is shown in pseudocode in Algorithm~\ref{alg:sorting}.   
% one can incrementally build a valid sorting of size $d$ by arbitrarily choosing valid move operations . all elements in $\mathrm{LCS}(p,q)$ into $\mathcal{T}$
%  incrementally updating the inverse permutation $p^{-1}$ of $p$. Note that   
 % Further, let $\mathcal{I}_p$ denote an array of length $m = \lvert p \rvert$ whose values $\mathcal{I}(\sigma)$ store the position of the element $\sigma$ in $p$. Assume also that $\mathcal{I}_p()$  
 %for every pair of elements $\sigma, \tau \in \mathcal{L}$, if $p_{\sigma} < p_{\tau}$ where $p_{\sigma}$ denotes the the position of $\sigma$ in $p$, then $q_{\sigma} < q_{\tau}$ must also be true. 
 %under certain insertion and deletion operations.

@@

\subsubsection{Proxy objective}\label{sec:proxy_objective}
When $d \sim O(m)$, the quadratic nature of the aforementioned greedy minimization can be computationally expensive. 
In order for Algorithm~\ref{alg:schedule} to be considered as an efficient alternative to vineyards---which requires $O(m \log m + k)$ preprocessing time to compute $k \sim O(m^2)$ transpositions---a solution which is subquadratic in $m$ is desirable. 
%In what follows, we present such an objective, showing its relation to the problem at hand as well its use in other known computational problems. 
% We show empirical results demonstrating its practical effectiveness as a heuristic in section~\ref{sec:results}.

@@

\section{Conclusion and Future Work}\label{sec:conclusion}
In conclusion, we presented a scheduling algorithm for efficiently updating a decomposition in coarse dynamic settings. Our approach is simple, easy to implement, and fully general: it does not depend on the geometry of underlying space nor the choice of triangulation. Moreover, we supplied efficient algorithms for our scheduling strategy, provided tight bounds where applicable, and demonstrated our algorithms performance with several real world use cases.
